Recursive Solutions to First-Order Model Counting

### 2022-08-16 doc/paper/paper.tex
Small changes

% TODO: FOMC algorithms -> of most algorithms to date
\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in finite-domain first-order logic. Despite
  being around for more than a decade, practical FOMC algorithms are still
  unable to compute functions as simple as a factorial. We argue that the
  capabilities of FOMC algorithms are severely limited by their inability to
  express arbitrary recursive computations. To enable arbitrary recursion, we
  relax the restrictions that typically accompany domain recursion and
  generalise circuits used to express a solution to an FOMC problem to graphs
  that may contain cycles. To this end, we enhance the most well-established
  (weighted) FOMC algorithm \textsc{ForcLift} with new compilation rules and an
  algorithm to check whether a recursive call is feasible. These improvements
  allow us to find efficient solutions to counting fundamental structures such
  as injections and bijections.
\end{abstract}
% TODO: We conclude after report on empirical results and discuss a few
% conjectures on what correctness might look like.

@@

% TODO: It might be worth remarking either in the abstract or in the intro that
% generally the weighted version is what is studied but the nonweighted version
% is also still interesting because of its connections to finite model theory.
% You could make a reference to the work of Paul Beam on symmetric weighted
% model counting.

@@

% in fact, the WFOMC algorithm by \citet{DBLP:conf/ijcai/BroeckTMDR11} can do so when it is unable to solve the instance in a lifted manner.

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \citep{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting \citep{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals \citep{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \citep{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\citep{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\citep{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \citep{DBLP:journals/ml/KimmigMG15} and statistical
relational \citep{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\citep{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.

@@

% TODO (footnote): You need to explain this a little bit more. You say that it
% is liftable in principle but you should clarif ythat factorial has never been
% shown empirically and there is no correctness theorem claiming that a
% factorial can be computed in a lifted manner. Or something like that.
However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{The problem of computing
  the factorial can be described using two variables and counting quantifiers,
  so it is known to be liftable in principle
  \citep{DBLP:journals/jair/Kuzelka21}.} We claim that this shortcoming is due
to the inability of these algorithms to construct recursive solutions. The topic
of recursion in the context of WFOMC has been studied before but in limited
ways. \Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data
that is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.
% TODO: Because this is the focus of your work you might want to give at least
% one example of such a stringent condition

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation.\footnote{There is also an
  intermediate stage called \emph{smoothing} that takes place between
  compilation and evaluation. As our changes to smoothing are rather elementary,
  we do not discuss them to not distract the reader from the main contributions
  of this paper.} In the first stage, various \emph{(compilation) rules} are
applied to the input (or some derivative) formula, gradually constructing a
circuit. In the second stage, the weights of the instance are propagated through
the circuit, computing the weighted model count. Along with new compilation
rules, \textsc{Crane} introduces changes to both stages of the process. First,
while \textsc{ForcLift} applies compilation rules via greedy\footnote{The
  algorithm is not described in any paper on \textsc{ForcLift} but can be found
  in its source code at \url{https://dtai.cs.kuleuven.be/drupal/wfomc}} search,
\textsc{Crane} uses a hybrid search algorithm that applies some rules greedily
and some using breadth-first search. Second, the product of compilation is not
directly evaluated but rather interpreted as a collection of functions on domain
sizes.

@@

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits, where cycles
represent recursive calls. Suppose the input formula $\phi$ depends on a domain
of size $n \in \mathbb{N}$. \emph{Generalised domain recursion} (GDR)---one of
the new compilation rules---transforms $\phi$ into a different formula $\psi$
that has an additional constant and some new \emph{constraints}. After some
additional transformations, the constraints in $\psi$ become `uniform' and can
be removed, replacing the domain of size $n$ with a new domain of size
$n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another compilation rule recognizes that the
resulting formula is just like the input formula $\phi$ except it refers to a
different domain. This observation allows us to add a cycle-forming edge to the
graph, which can be interpreted as function $f$ relying on $f(n-1)$ to compute
$f(n)$.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate sentences
in a variant of function-free many-sorted FOL with equality to this internal
format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all domains.
Initially, $\mathcal{D}$ contains all domains mentioned by the input (W)FOMC
instance. During compilation, new domains are added to $\mathcal{D}$ by some of
the compilation rules.
% TODO ('Let D... all domains'): You need to state explicitly or implicitly that
% to set of domains is constructed from a finite set of constants. Either you
% can define the set of constants in the langoage or you can say that in the
% interest of simplicity the set of constants is a set of nullary functions
% mentioned in the sentence, which only mentions relations or nullary functions.
% Function that take arguments are not allowed.

@@

% TODO: You should mention that somewhat unusually, first order variables are in
% capital letters, similar to the convention of logic programming. They are not
% to be confused with second order qualification.
One can read a formula from \cref{def:formula} as a sentence in a FOL\@. All
variables in a clause are implicitly universally quantified (but note that
variables are never shared among clauses), and all clauses in a formula are
implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit some parameters as well as nodes whose only arithmetic
    effect is multiplication by one. Some of these nodes play an important role
    in the weighted version of the problem, whereas others are remnants of the
    interaction between compilation rules and the way in which \textsc{ForcLift}
    handles existential quantifiers.}\label{fig:examplefcg}
\end{figure}

@@

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges.\footnote{Imposing an
  ordering on outgoing edges is just a limited version of edge labelling.} Node
labels consist of two parts: the \emph{type} and the \emph{parameters}. The type
of a node determines its out-degree.

@@

% Original domain recursion
\citet{DBLP:conf/nips/Broeck11} imposes stringent preconditions on the input
formula to ensure that the expanded version of the formula (as in
\cref{example:dr}) can be handled efficiently. The clauses in this expanded
formula are then partitioned into three parts based on whether the
transformation introduced constants or constraints or both. The aforementioned
conditions ensure that these parts can be treated independently.

@@

% TODO: explain notation (i.e., remove the bot)
\begin{example}
  Let $\phi = \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2')$, where
  \begin{align*}
    c_1' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

\paragraph{Hashing.}
We use (integer-valued) hash functions to efficiently discard pairs of formulas
that are too different for recursion to be established. The hash code of a
clause $c = (L, C, \delta_{c})$ (denoted by $\# c$) combines the hash codes of
the sets of constants and predicates in $c$, the numbers of positive and
negative literals, the number of inequality constraints $|C|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

@@

The compatibility of a pair of clauses is checked by function \texttt{genMaps}.
It considers every possible bijection $\beta\colon \Vars(c) \to \Vars(d)$ and
map $\gamma\colon \Doms(c) \to \Doms(d)$ s.t.
\[
  \begin{tikzcd}
    \Vars(c) \ar[r, dashed, "\beta"] \arrow[d, swap, "\delta_c"] & \Vars(d) \ar[d, "\delta_d"] \\
    \Doms(c) \ar[r, dashed, "\gamma"] \ar[d, hookrightarrow] & \Doms(d) \ar[d, hookrightarrow] \\
    \Doms(\psi) \ar[r, swap, "\rho", "|" marking, outer sep=5pt] & \Doms(\phi).
  \end{tikzcd}
\]
commutes, and $c$ becomes equal to $d$ when its variables are replaced according
to $\beta$ and its domains replaced according to $\gamma$. The function then
returns each such $\gamma$ as soon as possible. The commutativity check also
ensures that $\rho \cup \gamma$ is possible on \cref{line:recursion} of
\cref{alg:trycache}. The resulting (partial) function $\rho \cup \gamma$ is the
unique function s.t. $\rho \cup \gamma|_{\dom(\rho)} = \rho$, and
$\rho \cup \gamma|_{\dom(\gamma)} = \gamma$. The operation is defined when
$\rho|_{\dom(\rho)\cap\dom(\gamma)} = \gamma|_{\dom(\rho)\cap\dom(\gamma)}$.

@@

When \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11} compiles a WFOMC
instance into a circuit, each node type encodes an arithmetic operation on its
inputs and parameters. These operations are then immediately performed while
traversing the circuit and using domain sizes and weights as the initial inputs.
With \textsc{Crane}, the interpretation of an FCG is a collection of functions.
Each function has (some) domain sizes as parameters and may contain recursive
calls to other functions, including itself. While there may be any number of
subsidiary functions, there is always one main function that can be called with
the sizes of the domains of the input formula as arguments. Henceforth, this
function is always called $f$, and it is defined by the source node.

The interpretation of a node is decided by its type. Here we describe the
interpretations of new (or significantly changed) types and refer the reader to
previous work \citep{DBLP:conf/ijcai/BroeckTMDR11} for information on other
types. Both $\CR$ and $\GDR$ nodes do not contribute anything to the definitions
of functions---the interpretation of such a node is simply the interpretation of
its only direct successor. Obviously, $\star$ nodes also have no interpretation,
although for a different reason: incomplete FCGs are not meant to be
interpreted. The interpretation of a $\Reff$ node is a function call. The direct
successor of the $\Reff$ node (say, $v$) then must introduce a function. The
parameters of this function are the sizes of all domains used by nodes reachable
from $v$.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!10}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!10}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!10}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!10}\cmark & \xmark & & $\min\{\, m, n \,\}^2$ & --- & $mn$ \\
    \rowcolor{gray!10}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions.
    Here, $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that no complete solution was found.}\label{tbl:results}
\end{table*}
% TODO: This is very neat but readers from the community will expect to see some
% plots of time. Can you also give a few of those that reflect or mirror the
% conclusions from the table? Also it is a standard to discuss a few lifted
% theories like the smokers and friends example. Can you also include some
% empirical results with plots of time on such theories as well to appeal to the
% sanity of the reader that it actually works for problems in frameworks such
% as Markov log networks

>>>>

\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in finite-domain first-order logic. Despite
  being around for more than a decade, practical FOMC algorithms are still
  unable to compute functions as simple as a factorial. We argue that the
  capabilities of FOMC algorithms to date are limited by their inability to
  express arbitrary recursive computations. To enable arbitrary recursion, we
  relax the restrictions that typically accompany domain recursion and
  generalise circuits used to express a solution to an FOMC problem to graphs
  that may contain cycles. To this end, we enhance the most well-established
  (weighted) FOMC algorithm \textsc{ForcLift} with new compilation rules and an
  algorithm to check whether a recursive call is feasible. These improvements
  allow the algorithm to construct efficient solutions to counting fundamental
  structures such as injections and bijections. We end with a few conjectures on
  what classes of instances could be liftable as a result.
\end{abstract}

@@

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \citep{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting \citep{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals \citep{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \citep{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\citep{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\citep{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \citep{DBLP:journals/ml/KimmigMG15} and statistical
relational \citep{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\citep{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC tends to receive more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory \citep{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics \citep{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{The problem of computing
  the factorial can be described using two variables and counting quantifiers,
  so it is known to be liftable in principle
  \citep{DBLP:journals/jair/Kuzelka21}, but an algorithm that can lift such an
  instance in practice does not exist.} We claim that this shortcoming is due to
the inability of these algorithms to construct recursive solutions. The topic of
recursion in the context of WFOMC has been studied before but in limited ways.
\Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data that
is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy\footnote{The algorithm is not described in any paper but can be
  found in its source code at \url{https://dtai.cs.kuleuven.be/drupal/wfomc}}
search, \textsc{Crane} uses a hybrid search algorithm that applies some rules
greedily and some using breadth-first search. Second, the product of compilation
is not directly evaluated but rather interpreted as a collection of functions on
domain sizes.

% \footnote{There is also an intermediate stage called \emph{smoothing} that takes
%   place between compilation and evaluation. As our changes to smoothing are
%   rather elementary, we do not discuss them to not distract the reader from the
%   main contributions of this paper.}

@@

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits, where cycles
represent recursive calls. Suppose the input formula $\phi$ depends on a domain
of size $n \in \mathbb{N}$. \emph{Generalised domain recursion} (GDR)---one of
the new compilation rules---transforms $\phi$ into a different formula $\psi$
that has an additional constant and some new \emph{constraints}. After some
additional transformations, the constraints in $\psi$ become `uniform' and can
be removed, replacing the domain of size $n$ with a new domain of size
$n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another compilation rule recognizes that the
resulting formula is just like the input formula $\phi$ but with a different
domain. This observation allows us to add a cycle-forming edge to the graph,
which can be interpreted as function $f$ relying on $f(n-1)$ to compute $f(n)$.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate sentences
in a variant of function-free many-sorted FOL with equality to this internal
format. We use lowercase Latin letters for predicates and constants, uppercase
Latin letters for variables, and uppercase Greek letters for domains. An
\emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all (finite)
domains. Initially, $\mathcal{D}$ contains all domains mentioned by the input
(W)FOMC instance. During compilation, new domains are added to $\mathcal{D}$ by
some of the compilation rules. Each such new domain is interpreted as a subset
of some element of $\mathcal{D}$.

@@

One can read a formula from \cref{def:formula} as a sentence in a FOL\@. All
variables in a clause are implicitly universally quantified (but note that
variables are never shared among clauses), and all clauses in a formula are
implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit some parameters as well as nodes whose only arithmetic
    effect is multiplication by one.}\label{fig:examplefcg}
\end{figure}
% Some of these nodes play an important role in the weighted version of the
% problem, whereas others are remnants of the interaction between compilation
% rules and the way in which \textsc{ForcLift} handles existential quantifiers.

@@

% \footnote{Imposing an ordering on outgoing edges is just a limited version of
% edge labelling.}

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges. Node
labels consist of two parts: the \emph{type} and the \emph{parameters}. The type
of a node determines its out-degree.

@@

% Original domain recursion
\citet{DBLP:conf/nips/Broeck11} imposes stringent preconditions on the input
formula to ensure that the expanded version of the formula (as in
\cref{example:dr}) can be handled efficiently. For example, at least one clause
must contain a variable that occurs in all of its atoms. The clauses in this
expanded formula are then partitioned into three parts based on whether the
transformation introduced constants or constraints or both. The preconditions
ensure that these parts can be treated independently.

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2')$, where
  \begin{align*}
    c_1' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Delta \,\}),
           \end{multlined} \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

\paragraph{Hashing.}
We use (integer-valued) hash functions to discard pairs of formulas that are too
different for recursion to work. The hash code of a clause
$c = (L, C, \delta_{c})$ (denoted by $\# c$) combines the hash codes of the sets
of constants and predicates in $c$, the numbers of positive and negative
literals, the number of inequality constraints $|C|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

@@

Function \texttt{genMaps} checks the compatibility of a pair of clauses. It
considers every possible bijection $\beta\colon \Vars(c) \to \Vars(d)$ and map
$\gamma\colon \Doms(c) \to \Doms(d)$ s.t.
\[
  \begin{tikzcd}
    \Vars(c) \ar[r, dashed, "\beta"] \arrow[d, swap, "\delta_c"] & \Vars(d) \ar[d, "\delta_d"] \\
    \Doms(c) \ar[r, dashed, "\gamma"] \ar[d, hookrightarrow] & \Doms(d) \ar[d, hookrightarrow] \\
    \Doms(\psi) \ar[r, swap, "\rho", "|" marking, outer sep=5pt] & \Doms(\phi).
  \end{tikzcd}
\]
commutes, and $c$ becomes equal to $d$ when its variables are replaced according
to $\beta$ and its domains replaced according to $\gamma$. The function then
returns each such $\gamma$ as soon as possible. The commutativity check also
ensures that $\rho \cup \gamma$ is possible on \cref{line:recursion} of
\cref{alg:trycache}. The resulting (partial) function $\rho \cup \gamma$ is the
unique function s.t. $\rho \cup \gamma|_{\dom(\rho)} = \rho$, and
$\rho \cup \gamma|_{\dom(\gamma)} = \gamma$. The operation is defined when
$\rho|_{\dom(\rho)\cap\dom(\gamma)} = \gamma|_{\dom(\rho)\cap\dom(\gamma)}$.

@@

When \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11} compiles a WFOMC
instance into a circuit, each node type encodes an arithmetic operation on its
inputs and parameters. These operations are then immediately performed while
traversing the circuit. With \textsc{Crane}, the interpretation of an FCG is a
collection of functions. Each function has (some) domain sizes as parameters and
may contain recursive calls to other functions, including itself. While there
may be any number of subsidiary functions, there is always one main function
that can be called with the sizes of the domains of the input formula as
arguments. Henceforth, this function is always called $f$, and it is defined by
the source node.

The interpretation of a node is decided by its type. Here we describe the
interpretations of new (or significantly changed) types and refer the reader to
previous work \citep{DBLP:conf/ijcai/BroeckTMDR11} for information on other
types. The interpretation of a $\CR$ or a $\GDR$ node is simply the
interpretation of its only direct successor. Obviously, $\star$ nodes have no
interpretation as incomplete FCGs are not meant to be interpreted. The
interpretation of a $\Reff$ node is a function call. The direct successor of the
$\Reff$ node (say, $v$) then must introduce a function. The parameters of this
function are the sizes of all domains used by nodes reachable from $v$.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!10}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!10}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!10}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!10}\cmark & \xmark & & $\min\{\, m, n \,\}^2$ & --- & $mn$ \\
    \rowcolor{gray!10}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions.
    Here, $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that no complete solution was found.}\label{tbl:results}
\end{table*}
% TODO: It is a standard to discuss a few lifted theories like the smokers and
% friends example. Can you also include some empirical results on such theories
% as well to appeal to the sanity of the reader that it actually works for
% problems in frameworks such as Markov log networks

We compare \textsc{Crane} and \textsc{ForcLift}
\citep{DBLP:conf/ijcai/BroeckTMDR11} on their ability to count various kinds of
functions. This class of instances is chosen because of its simplicity and
ubiquity and the inability of state-of-the-art WFOMC algorithms to solve many
such instances. Note that other WFOMC algorithms---\textsc{L2C}
\citep{DBLP:conf/kr/KazemiP16} and probabilistic theorem proving
\citep{DBLP:journals/cacm/GogateD16}---are unable to solve any of the instances
that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.
